---
title: "Explaining suicide in Italian counties"
author: "Pietro Romozzi & Federico Colombo"
date: "20 novembre 2014"
output: html_document
bibliography:
- Reference2.bib
- RPackages.bib
- References3.bib
---

**Abstract:**

With the present work, the authors intend to create a descriptive model of economical, social and climatic characteristics of a territory, aiming to assess the impact of these factors on suicide rates. The model will be tested on Italian counties (110 observations), couniting on data obtained by ISTAT.


## Introduction:

### Topic

Several papers have tried to propose an explanatory framework to describe how climate conditions affect mental health. 
Climate may affect individuals’ mental health both directly through extreme climate change, which exposes people to trauma, 
or indirectly, as some climate’s characteristics may affect physical activity and mental health. Extreme temperature or 
humidity rates have an influence on individuals’ physical health as it becomes more energy consuming to do physical activity, 
to work or to travel, but also mental health could be negatively affected as climate erodes physical environments, which 
subsequently damage social environments and thus, mental condition. 
Health is defined by the World Health Organisation as a ‘state of complete physical, mental and social wellbeing and not 
merely the absence of disease or infirmity’ (World health organisation, 1948). Mental health is one of the key components 
that define general health and it can be seen as an alteration in thinking, mood, behaviour, and associated distress or impaired 
functioning. Usually, problems related to mental condition can be considered to be either uncommon (schizophrenia, anxiety and 
personal disorders) or common (depression, dementia, physiological distress).  The WHO showed that among almost 90% of people who 
died by suicide in high-income countries at least one mental disorder was present, while those affected by more than one mental 
disorder were experiencing significantly higher risks of committing suicide. 
The relationship between mental health (and the related suicide rates) and climate still needs to be studied into details, 
but some scholars identified extreme heat as a factor that is often associated with a general increase in aggressive, criminal 
and suicide rates (@heatwave). Humidity has also been associated with mental functioning, especially with poorer concentration 
and elevated fatigue. 
Climate change is a phenomenon which is expected to have severe consequences on the mental health of millions of people too: 
the threat to physical health or the potential loss of home, jobs and familiars due to catastrophic climate changes negatively 
affects mental condition. 
Sunlight is another factor, which has been documented to have direct effects on climate related health problems such as depressive 
symptoms, anxiety disorders, and other mental problems usually peaking in the cold season when there is limited sunlight (@Seasonaldisorder).
The debate about the impact of climate change on human health has only very recently included considerations of mental health. 
Climate can have an influence on our psyche in a varieties of ways: it can directly inflict more and more natural disasters on 
human settlements, which, as evidence suggests, increases anxiety-related or chronic severe mental problems; climate can also 
increase the risk of physical injuries or health problems, which are correlated with mental health; finally, climate can endanger 
the natural and social environment on which people depend for their wellbeing.


### Relevance

One obvious reason why this work is potentially interesting is that mental disorders and suicide are nowadays often discussed and debated both among insitution in health's field and media.
Indeed, according to the latest WHO report concerning suicide prevention [see the latest report by @PreventingSuicideWHO], "Globally, among young adults 15-29 years of age suicide accounts for 8.5% of all deaths and is ranked as the second leading cause of death (after traffic accidents). Among adults aged 30-49 years it accounts for 4.1% of all deaths and is ranked the fifth leading 
cause of death". This statement displays clearly the relevance of this issue, which is also greater considering that these estimation are often underrated because of the difficulties in recognizing sucide as a cause of death - will of committing suicide could be not recognized, families could prefer not to speak about it, and so on [see the note by @NotaIstat2012].
The above mentioned paper, moreover, is considered by several OECD countries as a strong stimulus to move further in suicide prevention.
Among these countries, we are going to focus on Italy, particularly at provincial level. 
In particular, our convintion is that a focus on smaller-than-countries environments, does help in assessing how factors affecting suicides impact differently within the same country, thus highlighting a possible variance among regions which is overlooked by country-level aggregate data.
In addition to this "micro-level" focus, we also intend to stress some environmental and climatic characteristics which could contribute to create "suicide-genic" environment.
In fact, according to a recently developed body of studies, there is a correlation between environmental-climatic characteristics and mental health. So, considering mental disorder as a possible driver to suicide, we include these types of variables in our study trying to widen the range of explanatory factors for the phenomenon.
The above mentioned link is often not considered, as some auhors like @Berry2009 underline, therefore we believe that including them in the present paper could deliver interesting results and perhaps stimulate further research in this branch.
We also do not neglet the use of more "ortodox" variables (i.e. related to economic and social dimensions) to develop a fairly comprehensive perspective.
Looking for empirical evidence from Italian data, our aim is to add knowledge to existing literature reporting specific suicide patterns in Italy, among which we will look for evidence of the magnitude of climatic and environmental factors, believing that our wide-spectrum approach makes the present work interesting and innovative.
Another interesting feature is the provincial focus which allows to underscore, in the end, contingent specific characteristics of the territory rather than the whole country. 
Recognizing these characteristics, will hopefully provide information to policy makers for setting efficient preventive strategies at a more manageable and flexible micro-level, serving as an input for tailored and focused polices.


## Research questions:

As said above, this paper will focus on the relation between suicidal behaviours (an extreme consequence of mental disorders) and various factors concerning climate [@Seasonaldisorder] and socio-economic context.

Regarding *climate*, we want to test if the following hypoteses hold true:

- Suicide rate in a certain Italian county increases when the average temperature is low [@heatwave].

- Suicide rate in a certain Italian county increases when the average temperature range is high

- Suicide rate increases when precipitations happen to be frequent.

Regarding the *socio-economical framework* instead, we want to verify the follwing hypoteses:

- Suicide rate increases when GDP per capita is high [@AGlance2013]. 

- Suicide rate increases according to economic inequality (represented by the gini index).

- Suicide rate increases whereby unemployment rate is higher.

- Suicide rate is higher in more populated counties.

The hypotheses we want to test originate from a careful review of the relevant literature regarding mental disorders, suicidal patterns and the magnitude of climate impact on them.


## Methodology:

### Sources and data gathering

For the scope of this work, data for different variables were necessary for all of the Italian counties; in particular we looked for data about temperature, temperature range, precipitations, GDP, gini's index, unemployment, population, and naturally suicides which is our depending variable.
We decided to carry out our analysis on a single year, because of the higher complexity of panel data both in analysis phase and in data gathering process. To have a significant result, we believed that recent data were necessary, so we identified 2009 as an acceptable compromise between proximity in time and availability of data for all of our variables, thus all data gathered and used regard 2009.
Most of the data required were founded in the Italian National Institute of Statistics (ISTAT) database, and fortunately the missing ones were available in other databases of ISTAT-related agencies or in the Eurostat database.
Accessing to the ISTAT database is free and easy, and data are clear and well structured. 
Nevertheless, we needed to process them in order to obtain so-called **tidy data**, required by R to analyze and modeling them. We recall a definition of tidy data, according to [@Wickham2014]:

1. Each variable forms a column.

2. Each observation forms a row.

3. Each type of observational unit forms a table.

Given that, we firstly collected and downloaded all the data from ISTAT website (**http://www.istat.it/it/**) and from ISTAT-related agencies' databases. Then, considering that the amount of observation was "human-manageable", for practical reasons we carried out the tidying process in Microsoft Excel.
Basically, we made sure that all names of different counties were the same in all datasets we downloaded (otherwise we simply renamed the counties), as well as checking that our values were expressed in the appropriate scale, further discussed below. For istance, we decided to rescale GDP per capita from Euro to thousands Euros, to facilitate the interpretation of our results.
To sum up, we downloaded data and tidied them usinc Excel obtaining three dataframes, namely **gdp_data**, **socio_eco_data** and **climate_data**. Then we converted them in .csv format, easily readable by R, and finally we moved into RStudio to merge these three dataframes.
In the end we collected 110 observations, one for each Italian county, of 9 variables.

### Considered variables and analysis structure:

The variable *suicide rate* is defined by the WHO as the standardized number of deaths by suicide every 100.000 people occurred in a certain country. 
We collected data about suicide rate for every Italian county, from the ISTAT database.

We adopted the variable *population* as a proxy of the degree of urbanization of the area. It describes the total number of people living in the county. 
Carrying out our analysis, we decided to take value for population in logaritmic scale.
Also in this case, we found the data about population in the ISTAT database.

Regarding the variable *GDP per capita*, we found in the ISTAT database a precious source. Nevertheless, the data provided were expressed in Euro, deemed not to provide significant impact in our linear analysis since a one-euro variation is certainly too small to impact on suicide rate. Therefore, using Excel during our data cleaning process, we expressed GDP per capita in thousands Euro, so that our coefficient will represent the impact on suicide rate caused by a variation of 1000 Euro, perhaps delivering more significant results.

The *Gini Coefficient*, is the most common measure of income inequality. It varies between 0 and 1 and expresses inequality among values of the frequency distribution of income for our target counties. Even these data were available via ISTAT, which provides the index either including households' income from the rent of owned properties or not. Since in Italy, has been culturally widespread to invest in real estate, at least until the economic crisis of 2008 and the subsequent introduction of a strong tax levied on real estate properties, we believe that adopting the Gini index including rent might better represent the italian context.
As expected, we also notice gini's index to be slightly higher in Southern counties compared to the Centre and the North.

We included in our analysis also *unemployment rate* since economical, social and psychological consequences of job loss might have an influence in mental disorders and eventually in suicidal behavior.
This data, provided by ISTAT refers to the share of the labor force that is without work but available for and seeking employment.

The variable *average temperature* simply corresponds to the value of the annual average temperature in a certain county expressed in Celsius degrees. Insted, *average temperature range* corresponds to the average difference between the maximum and the minimum temperature registered in the county, again expressed in Celsius degrees. Then, *precipitations* measures the annual quantitiy of rain expressed in millimiters. Those data about climate are obtained from CRA-CMA, a ISTAT-related agency, particularly the department for meteorology applied to agriculture [@NotaIstat2012].

Lastly, we would like to include also *solar radiation* measuring the level of solar irradiance in megajoul (MJ) per square meter. The Italian Air Force offers a metereological service, which is responsible for the measurement of solar irradiance on the Italian territory. Although we intended to use this variable, county-level data were not available, so eventually we dropped it.

We believe that a linear model was an effective way to analyze our dataframe because it could give a clear idea of the impact of the selected indipendent variables on suicides which we chose as to be the dependent variable.
As the reader will observe below, we started from a linear model including all the variables we considered, then recognizing that some variables were particularly unsignificant, we opted to remove two of them obtaining a simplified model which appeared fairly significant and interesting.
In the end we carried out our analysis providing two linear models, whose interpretation will be discussed further, then we enrich the statistical output with some visual displays, particularly focusing on the relationship between suicides and climatic factors.

All the data gathered will be processed using "R" [see @CiteR]. We use as well some packages developed for R, namely **ggplot2** [see @R-ggplot2], **stargazer** [see @stargazer], **repmis** [see @R-repmis], **car** [see @Rcar] and **knitr** [see @knitr].


### Case studies:

In a previous work, we tried to test the model we will use in this analysis on regional data. 
However we faced the big issue of not having enough observations, given that Italy counts only 20 regions and it resulted in an unsignificant model. 
Nevertheless, we were fairly confident in the goodness of our model, therefore we decided to test it again on a sufficient number of observations. This led us to move our analysis to a county-level, thus having available 110 observations.
Eventually, in the present work we take into consideration all the Italian counties.

```{r, include=FALSE, echo=FALSE}

# WD to be set before starting.

# Loading tidy data frames that we have to merge.

meteo <- read.csv("climate_data.csv", header = TRUE, sep = ";", dec = ",",fill = TRUE)

gdp <- read.csv("gdp_data.csv", header = T, sep =";", dec = ",", fill = T)

main <- read.csv("socio_eco_data.csv", header = T, sep =";", dec = ",", fill = T)

# Merging

Merging1 <- merge(x = main , y = meteo , by = "county", all = T )

Dataset <- merge(x = Merging1, y = gdp, by = "county", all = T )

```

## Descriptive Statistics: 

```{r, results = 'asis', echo = FALSE, message=FALSE}

# Descriptive statistics

# Summary for the variables used
library(stargazer)

stargazer(Dataset, summary = T, title = "Data Description", digits = 2, type = "html", header = F)

# Scatterplot matrix
require("car")

car :: scatterplotMatrix(Dataset[2:9], main = "Scatterplot Matrix")
```

## Proposed Model:

The model we want to estimate will therefore be the following: 

$$
suicide_rate = \alpha + \beta_{1}GDP_per_capita + \beta_{2}unemployment_rate + 
\beta_{3}population + \beta_{4}gini_index + \beta_{5}average_temperature + \beta_{6}average_temperature_range + \beta_{7}average_precipitations + e
$$

### A first linear model using all the explanatory variables:
```{r, results='asis', echo=FALSE, message=FALSE}

# Our Model

LinModel1 <- lm(st_suicide_rate ~ gdp_per_capita + unemployment_rate + log(population) + gini_index + avg_temperature + avg_temp_range + avg_precipitations, data = Dataset)

# Summary Table of the Model

stargazer(LinModel1, title = "Estimation of the First Model", digits = 3, type = "html", header = F)

```

### A second, simplified, model:

$$
SuicideRate = \alpha + \beta_{1} + \beta_{2}UnemploymentRate + 
\beta_{3}population + \beta_{4}average_temperature + \beta_{5}average_precipitations + e
$$

```{r, results='asis', echo=FALSE, message=FALSE}

# Dropped avg_temp_range and gini_index

LinModel2 <- lm(st_suicide_rate ~ gdp_per_capita + unemployment_rate + log(population) + avg_temperature + avg_precipitations, data = Dataset)

stargazer(LinModel2, title = "Estimation of the Second Model", digits = 3, type = "html")

```

```{r, results='asis', echo=FALSE, message=FALSE}

stargazer(LinModel1, LinModel2, title = "Models Comparison", type = "html", digits = 3, align = T)

```

```{r, results ='asis', echo = FALSE, message=FALSE}

CoefTab <- coefficients(LinModel2)

knitr :: kable(CoefTab, format = "html", digits = 3, caption = "Table of Coefficients")

Confint <- confint(LinModel2)

knitr :: kable(Confint, format = "html", digits = 3, caption = "Table of Coefficients")

```

```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE}

plot(LinModel2, which = 1)

plot(LinModel2, which = 2)

library(ggplot2)

ggplot2::ggplot(data = Dataset, aes(x = avg_precipitations, y = st_suicide_rate), message = F) + geom_point() + stat_smooth() + theme_bw() + ggtitle("Suicide Rate vs Precipitation in Italy")

```


## Main findings and discussion:


## Conclusion:

## References:

