---
title: "Explaining suicide in Italian counties"
author: "Pietro Romozzi & Federico Colombo-Ercole"
date: "02 dicembre 2014"
output:
  ioslides_presentation:
    css: http://maxcdn.bootstrapcdn.com/font-awesome/4.2.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png

---

# What and Why

## A quick background: {.flexbox .vcenter}

- Among young adults (15-29), suicide accounts for 8.5% of all deaths: *second leading cause of death* (after traffic accidents). 
- Among adults (30-49) suicide accounts for 4.1% of all deaths: *fifth leading cause of death* (latest "Prevent Suicide" report by WHO).

- "The debate about the impact of climate change on human health has, very recently, included consideration of mental health" (Berry, Bowen & Kjellstrom, 2009).

- The *mental health vs climate* relationship still needs to be studied into details (**research gap**).

## Research question: {.flexbox .vcenter}

*How do climate factors affect suicide rate in the Italian context?*

- Is suicide rate lower whereby the average temperature is higher?
- Is suicide rate higher whereby the average temperature range is higher?
- Is suicide rate higher whereby precipitations happen more frequently?

# Methodology

## Case selection: {.smaller}

Previous study: regional data, 20 observations, unsignificant model.

Now, Italy is still our focus: *county-level data* provide 110 observations!

```{r, echo=FALSE, message=FALSE, warning=FALSE}

library(ggmap)

location <- "Italy"

qmap(location = "Roma", zoom = 6, maptype = "hybrid")

```


## Our Model: {.flexbox .vcenter} 

$$
<<<<<<< Updated upstream
StdSuicideRate = \alpha + \beta_{1}GdpPerCapita + \beta_{2}Unemployment
=======
SuicideRate = \alpha + \beta_{1}GdpPerCapita + \beta_{2}UnemploymentRate
>>>>>>> Stashed changes
$$
$$
+ \beta_{3}Population + \beta_{4}GiniIndex + \beta_{5}AverageTemperature 
$$
$$
+ \beta_{6}AverageTemperatureRange + \beta_{7}AveragePrecipitations + e
$$

## A Second, improved, model: {.flexbox .vcenter}


$$
<<<<<<< Updated upstream
StdSuicideRate = \alpha + \beta_{1}GdpPerCapita + \beta_{2}Population
$$
$$
+ \beta_{3}AverageTemperature + \beta_{4}AveragePrecipitations + e 
$$

=======
SuicideRate = \alpha + \beta_{1}GdpPerCapita + \beta_{2}UnemploymentRate 
$$
$$
+ \beta_{3}AveragePrecipitations + \beta_{4}AverageTemperature 
$$
$$
+ \beta_{5}Population + e
$$


--------
>>>>>>> Stashed changes

```{r, echo=FALSE, results='asis'}

# Loading tidy data frames that we have to merge.

meteo <- read.csv("climate_data.csv", header = TRUE, sep = ";", dec = ",",fill = TRUE)

gdp <- read.csv("gdp_data.csv", header = T, sep =";", dec = ",", fill = T)

main <- read.csv("socio_eco_data.csv", header = T, sep =";", dec = ",", fill = T)
  
# Merging

Merging1 <- merge(x = main , y = meteo , by = "county", all = T )

Dataset <- merge(x = Merging1, y = gdp, by = "county", all = T )
```

# Results

## Models Comparison: {.flexbox .vcenter}


First Model           | Second Model          |
--------------------- | ----------------------|
GDP per capita***     | GDP per capita***     |
Unemployment Rate     | *dropped*             |
Population***         | Population***         |
Gini index            | *dropped*             |
Avg Temperature       | Avg Temperature       |
Avg Temp. Range       | *dropped*             |
Avg Precipitations*** | Avg Precipitations*** |


## Findings: {.flexbox .vcenter}

- **Average temperatures** and **Temperature range** have no impact on suicide rate across Italy 

- **GDP per capita** has a positive impact on suicides (in Italy).

- **Population size** has a negative impact on suicides.

- **Annual precipitations** has a negative impact on suicides.


## Predicted Probabilities:

```{r,message=FALSE,echo=FALSE,results='asis', warning=FALSE}

library(Zelig)

CompleteDataset <- na.omit(Dataset)

Z1 <- zelig (st_suicide_rate ~ gdp_per_capita + log(population) + avg_temperature + avg_precipitations, cite = F, data = CompleteDataset, model = "ls")

x.out2 <- setx(Z1, avg_precipitations = 700:1200)

simz2 <- sim(Z1, x = x.out2)

plot(simz2, xlab = 'Average precipitations in mm, year 2009')

```

---------

```{r,message=FALSE,echo=FALSE,results='asis', warning=FALSE}
library(ggplot2)

LinModel2 <- lm(log(st_suicide_rate) ~ log(gdp_per_capita) + log(population) + log(avg_temperature) + log(avg_precipitations), data = Dataset)

ggplot2::ggplot(data = Dataset, aes(x = avg_precipitations, y = log(st_suicide_rate), message = F)) + geom_point() + stat_smooth() + theme_bw() + ggtitle("Suicide Rate vs Precipitation in Italy")
```

## Conclusion: {.flexbox .vcenter}

- *Climate characteristics do affect suicide rates in Italy.*

- Correlation not as strong as (we) would have expected.

- Rain actually has a negative impact on suicidial behaviour (contrary to what is commonly believed).

- A one percentage increase in average precipitations decreases
suicide rate by 0,461 percent.

## Limitations and Further Research: {.flexbox .vcenter}

- Data: only 2009. Panel data analysis?

- Further applications to test whether precipitations and sucide rates are negatively correlated also in other developed countries.

- Test other climatic factors (e.g. solar radiation)?

- Test it with monthly data looking for seasonal patterns for suicide.

# Thank you for your attention! {.flexbox .vcenter}
